The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-06 Fsing/FThresholdsPaper/FThresholds.tex
Edits to introduction of FThresholds paper

Recall that a ring of prime characteristic $p>0$ is said to be \emph{$F$-pure} if the Frobenius map | that is, ring endomorphism sending an element to its $p$-th power | is a pure morphism;  under natural geometric hypotheses, this is equivalent to the condition that the Frobenius morphism splits as a map of rings.

@@

After the advent of tight closure \cite{HochsterHunekeTC1}, the usage of Frobenius to quantify singularity | that is, deviation from regularity | proliferated, and based on a connection discovered between $F$-pure and \emph{log canonical} singularities \cite{HaraWatanabeFRegFPure}, the concept of $F$-purity was generalized to the context of \emph{pairs}.
% $(R, f^t)$, where $f$ is an element of $R$, and $t$ is a nonnegative real number considered a formal exponent.

@@

Unfortunately, the $F$-pure threshold is typically difficult to calculate, even in many concrete cases.   It is also not so well-understood from the perspective of reduction to prime characteristic.   Nevertheless, $F$-pure thresholds have been the focus of intense study over the past fifteen years, and the \emph{Macaulay2} package \emph{FThresholds} is centered around calculating and estimating the $F$-pure threshold and other $F$-thresholds, with the function {\tt fpt} at its core.

\subsection*{Some background and notation}  For most of article, we work in a polynomial ring over a finite field of  characteristic $p>0$.  Unless otherwise stated, all polynomials and ideals under consideration are regarded as elements in this ambient ring.  The ideal of this ring generated by the ambient variables is denoted $\idealm$.

Let us roughly outline how many natural numerical invariants in prime characteristic commutative algebra are constructed:  For every natural number $e$, associate to some fixed set of data | often, a collection of polynomials and ideals | an integer describing something of relevance (e.g., the dimension of some interesting vector space constructed in terms of your initial data).   Finally, normalize this integer by dividing by some power of $p^e$, and then take the limit as $e \to \infty$.

Consider a nonzero polynomial $f$ and a natural number $e$.  If $f$ does not vanish at the origin, then set set $\nu_f^{\idealm}(p^e) := \infty$.  Otherwise, $f \in \idealm$, and we instead define
\[ \nu_f^{\idealm}(p^e) := \max \{ n : f^n \notin \idealm^{[p^e]} \} \] 
where $\idealm^{[p^e]}$ denotes the $p^e$-th Frobenius power of $\idealm$, that is, the ideal generated by the $p^e$-th powers of the elements of $\idealm$.    It is not very hard to see that this number is exactly the largest power of $f$ that is part of a free basis for the localization of the ambient polynomial ring at the origin over its subring of $p^e$-th powers.  As noted earlier, this number also has a natural interpretation in terms of $F$-purity of pairs.

@@

Inspired by the connections with $F$-purity of pairs,  this limit is called the \emph{$F$-pure threshold} of $f$ at the origin.  This limit exists, and is finite when $f \in \idealm$.  In the latter case, the limit is also rational, but this is far from obvious  \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds}.

The $F$-pure threshold is closely related to a many other fundamental concepts in prime characteristic commutative algebra.   For instance, 
\[ \ft^{\idealm}(f) = \inf \{ t : \tau(f^t)  \subseteq \idealm \} = \sup \{ t : \sigma(R, f^t) \neq 0 \} \]
where $\tau(f^t)$ and $\sigma(f^t)$ are the \emph{test ideal} and \emph{$F$-signature}, respectively, associated to $f$, and the formal nonnegative real exponent $t$.  The former is an ideal in the ambient ring of $f$, and the latter is a real number;  both depend on the parameter $t$ and prime $p$ in subtle ways \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds, BlickleSchwedeTuckerFSigPairs1}.
Though we will not elaborate much on these topics,  we stress that the \emph{Macaulay2} package {\tt TestIdeals} provides functionality for the effective computation of test ideals in many cases \cite{TestIdealsPackage, TestIdealsPaper}.

In the literature, the $F$-pure threshold $\ft^{\idealm}(f)$ is often denoted $\fpt(f)$, for obvious reasons.  However, in this note, we adopt the former notation to avoid any possible confusion with the function {\tt fpt} described in Section \ref{sec.FPT}.  Indeed, the command {\tt fpt(f)} need not always output the number $\ft^{\idealm}(f) = \fpt(f)$, though in the case that it cannot, it will provide a reasonably good estimate of this number.

It turns out that the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ itself, and not just its limit, also encodes interesting information about $f$.  For example, it is closely related to the so-called \emph{Bernstein-Sato polynomial} \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato}.  It is then natural to ask whether one can recover the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ from the limit $\ft^{\idealm}(f)$.  Fortunately, the answer to this question is yes \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato, HernandezFPurityOfHypersurfaces}. More precisely, 
\begin{equation}
\label{sequence-from-limit}
\nu_f^{\idealm}(p^e) = \lceil p^e \cdot \ft^{\idealm}(f) \rceil - 1.  
\end{equation}

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FThresholds}.
\begin{itemize}
\item \autoref{sec.Nu} describes the methods implemented for approximating the $F$-pure threshold and other $F$-thresholds.  
\item \autoref{sec.IsFPT} explains the package's functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}. 
\end{itemize}

@@

\subsection*{Acknowledgements.}  First and foremost, the authors enthusiastically thank all authors and contributors to the \emph{FThresholds} package: (INSERT NAMES HERE). %Erin Bela, Juliette Bruce, Zhibek Kadyrsizova, Sara Malec, Maral Mostafazadehfard, Marcus Robinson, Dan Smolkin, and Robert Walker. 
We are also grateful the organizers of the \emph{Macaulay2} workshops where much of the functionality described herein was developed, hosted by Wake Forest University in 2012, the University of California, Berkeley in 2014 and 2017, Boise State University in 2015, and the University of Utah in 2016. 
We are also grateful to the University of Utah for hosting a collaborative development visit in 2018, and to the Institute of Mathematics and its Applications for its support for the Coding Sprint \emph{F-thresholds in Macaulay2} in 2019, where the current version of the package was finalized and this paper was completed.

>>>>

Recall that a ring of prime characteristic $p>0$ is said to be \emph{$F$-pure} if the Frobenius map|that is, the ring endomorphism sending an element to its $p$-th power|is a pure morphism;  under natural geometric hypotheses, this is equivalent to the condition that the Frobenius morphism splits as a map of rings.

@@

After the advent of tight closure \cite{HochsterHunekeTC1}, the usage of Frobenius to quantify singularity|that is, deviation from regularity|proliferated, and based on a connection discovered between $F$-pure and \emph{log canonical} singularities \cite{HaraWatanabeFRegFPure}, the concept of $F$-purity was generalized to the context of \emph{pairs}.
% $(R, f^t)$, where $f$ is an element of $R$, and $t$ is a nonnegative real number considered a formal exponent.

@@

Unfortunately, the $F$-pure threshold is typically difficult to calculate, even in many concrete cases.   It is also not so well understood from the perspective of reduction to prime characteristic.   Nevertheless, $F$-pure thresholds have been the focus of intense study over the past fifteen years, and the \emph{Macaulay2} package \emph{FrobeniusThresholds} is centered around calculating and estimating the $F$-pure threshold and other $F$-thresholds, with the function {\tt fpt} at its core.

\subsection*{Some background and notation}  For most of this article, we work in a polynomial ring over a finite field of  characteristic $p>0$.  Unless otherwise stated, all polynomials and ideals under consideration are regarded as elements in this ambient ring.  The ideal of this ring generated by the ring's variables is denoted $\idealm$.

Let us roughly outline how many natural numerical invariants in prime characteristic commutative algebra are constructed:  For every natural number $e$, associate to some fixed set of data|often, a collection of polynomials and ideals|an integer describing something of relevance that depends on $e$ (e.g., the dimension of some interesting vector space constructed in terms of your initial data).   Finally, normalize this integer by dividing by some power of $p^e$, and then take the limit as $e \to \infty$.

Consider a nonzero polynomial $f$ and a natural number $e$.  If $f$ does not vanish at the origin, then set set $\nu_f^{\idealm}(p^e) := \infty$.  Otherwise, $f \in \idealm$, and we instead define
\[ \nu_f^{\idealm}(p^e) := \max \{ n>0 : f^n \notin \idealm^{[p^e]} \} \] 
where $\idealm^{[p^e]}$ denotes the $p^e$-th Frobenius power of $\idealm$, that is, the ideal generated by the $p^e$-th powers of the elements of $\idealm$.    It is not very hard to see that this number is exactly the largest power of $f$ that is part of a free basis for the localization of the ambient polynomial ring at the origin over its subring of $p^e$-th powers.  As noted earlier, this number also has a natural interpretation in terms of $F$-purity of pairs.

@@

Inspired by its connections with the $F$-purity of pairs,  this limit is called the \emph{$F$-pure threshold} of $f$ at the origin.  This limit exists, and is finite when $f \in \idealm$.  In the latter case, the limit is also rational, but this is far from obvious  \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds}.

The $F$-pure threshold is closely related to many other fundamental concepts in prime characteristic commutative algebra.   For instance, 
\[ \ft^{\idealm}(f) = \inf \{ t>0 : \tau(f^t)  \subseteq \idealm \} = \sup \{ t>0 : \sigma( f^t) \neq 0 \} \]
where $\tau(f^t)$ and $\sigma(f^t)$ are the \emph{test ideal} and \emph{$F$-signature}, respectively, associated to $f$ and the formal nonnegative real exponent $t$.  The former is an ideal in the ambient ring of $f$, and the latter is a real number;  both depend on the parameter $t$ and prime $p$ in subtle ways \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds, BlickleSchwedeTuckerFSigPairs1}.
Though we will not elaborate much on these topics,  we stress that the \emph{Macaulay2} package \emph{TestIdeals} provides functionality for the effective computation of test ideals in many cases \cite{TestIdealsPackage, TestIdealsPaper}.

In the literature, the $F$-pure threshold $\ft^{\idealm}(f)$ is often denoted $\fpt(f)$, for obvious reasons.  However, in this note, we adopt the former notation to avoid any possible confusion with the \emph{Macaulay2} function {\tt fpt} described in Section \ref{sec.FPT}.  Indeed, the command {\tt fpt(f)} need not always output the number $\ft^{\idealm}(f) = \fpt(f)$, though in the case that it cannot, it will provide a reasonably good estimate of this number.

It turns out that the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ itself, and not just its limit, also encodes interesting information about $f$.  For example, it is closely related to the so-called \emph{Bernstein-Sato polynomial} \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato}.  Thus, it is natural to ask whether one can recover the sequence $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ from the limit $\ft^{\idealm}(f)$.  Fortunately, the answer to this question is yes \cite{MustataTakagiWatanabeFThresholdsAndBernsteinSato, HernandezFPurityOfHypersurfaces}. More precisely, 
\begin{equation}
\label{sequence-from-limit}
\nu_f^{\idealm}(p^e) = \lceil p^e \cdot \ft^{\idealm}(f) \rceil - 1.  
\end{equation}

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item \autoref{sec.Nu} describes the methods implemented for approximating the $F$-pure threshold and other $F$-thresholds.  
\item \autoref{sec.IsFPT} explains the package's functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}. 
\end{itemize}

@@

\subsection*{Acknowledgements.}  First and foremost, the authors enthusiastically thank all authors and contributors to the \emph{FrobeniusThresholds} package.
Our package coauthors are (LIST NAMES), and others who contributed code are (LIST OTHER NAMES). %Erin Bela, Juliette Bruce, Zhibek Kadyrsizova, Sara Malec, Maral Mostafazadehfard, Marcus Robinson, Dan Smolkin, and Robert Walker. 
We are also grateful the organizers of the \emph{Macaulay2} workshops where much of the functionality described herein was developed, hosted by Wake Forest University in 2012, the University of California, Berkeley in 2014 and 2017, Boise State University in 2015, and the University of Utah in 2016. 
We are also grateful to the University of Utah for hosting a collaborative development visit in 2018, and to the Institute of Mathematics and its Applications for its support for the Coding Sprint \emph{F-thresholds in Macaulay2} in 2019, where the current version of the package was finalized and this paper was completed.
